---
title: Progress on svy2lme
author: Thomas Lumley
date: '2018-10-19'
slug: progress-on-svy2lme
categories: []
tags: []
---

The [`svy2lme`](https://github.com/tslumley/svylme) package for linear mixed models under complex sampling may still contain nuts, but at least the user interface has settled down and it gives plausible answers for some toy examples. The recent change is to compute pairwise sampling probabilities from a survey design object, rather than some horrible set of separate specifications.  It still doesn't support complicated PPS designs, but since the survey package does, that should be feasible.

Here's a toy example, with comparison to Stata's `mixed` function using nested pseudolikelihood. First, simulate some data.  We create a population with 1000 clusters of size 20, then take a biased sample of 177 of the clusters and a simple random sample of 5 observations from each cluster.

```{r}
set.seed(2000-2-29)
df<-data.frame(x=rnorm(1000*20),g=rep(1:1000,each=20), t=rep(1:20,1000), id=1:20000)
df$u<-with(df, rnorm(1000)[g])
df$y<-with(df, x+u+rnorm(1000,s=2))

## oversample extreme `u` to bias random-intercept variance
pg<-exp(abs(df$u/2)-2.2)[df$t==1]
in1<-rbinom(1000,1,pg)==1
in2<-rep(1:5, length(in1))
sdf<-subset(df, (g %in% (1:1000)[in1]) & (t %in% in2))

p1<-rep(pg[in1],each=5)
N2<-rep(20,nrow(sdf))

sdf$w1<-1/p1
sdf$w2<-20/5
 
design<-survey::svydesign(id=~g+id, data=sdf, weights=~w1+w2)
```

Fitting the linear mixed model looks like it does using `lmer` 
```{r}
pair<-svylme::svy2lme(y~x+(1|g),design=design)
pair
```

Now, in Stata: the command is

```
mixed y x [pw=w2]|| g:,pweight(w1) pwscale(gk)
```

giving output

```
                                                Wald chi2(1)      =     173.51
Log pseudolikelihood = -44980.757               Prob > chi2       =     0.0000

                                    (Std. Err. adjusted for 177 clusters in g)
------------------------------------------------------------------------------
             |               Robust
           y |      Coef.   Std. Err.      z    P>|z|     [95% Conf. Interval]
-------------+----------------------------------------------------------------
           x |   1.046958   .0794822    13.17   0.000     .8911759     1.20274
       _cons |   -.075619    .098788    -0.77   0.444    -.2692398    .1180019
------------------------------------------------------------------------------

------------------------------------------------------------------------------
                             |               Robust           
  Random-effects Parameters  |   Estimate   Std. Err.     [95% Conf. Interval]
-----------------------------+------------------------------------------------
g: Identity                  |
                  var(_cons) |   1.034899   .1671189      .7541251    1.420211
-----------------------------+------------------------------------------------
               var(Residual) |   3.728113   .2021218      3.352283    4.146078
------------------------------------------------------------------------------
```

So you don't have to do square roots yourself, here are the R variance components on the variance scale

```{r}
coef(pair, random=TRUE)
```


And, for comparison, an unweighted model: the sampling is doing something

```{r}
lme4::lmer(y~x+(1|g),data=sdf)
```


It would be really nice to have a proper example, but the ones I can find in the literature are like *if you wade through all these files and extract these variables and do some additional calculations you'll get the data we used*. And, you know, eventually I probably will, but I'm still hoping for something more concrete.