---
title: svycontrast
author: Thomas Lumley
date: '2018-12-10'
slug: svycontrast
categories: []
tags: []
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE, warning=FALSE)
```


I got asked for more detail about the `svycontrast()` function, so I thought I'd post it here too.  The function is related to the `CONTRASTS` you get in SAS, but focused on estimation rather than testing. 

The input to `svycontrast()` is a $p$-vector of estimates $\hat\theta$ (which I'll consider as a column vector) and an estimated $p\times p$ covariance matrix $\hat\Xi$

There are two main cases:

### Linear
Given a  $p$-vector of coefficients $b$, the function computes $b^T\hat\theta$ and $b^T\hat\Xi b$.  Given a list of $k$ $p$-vectors of coefficients the function pastes them into a $p\times k$ matrix $B$ and computes $B^T\theta$ and $B^T\hat\Xi B$. 

For example, from the help page
```{r}
library(survey)
data(api)
dclus1<-svydesign(id=~dnum, weights=~pw, data=apiclus1, fpc=~fpc)
a <- svymean(~api00+enroll+api99, dclus1)
b<-svycontrast(a, list(avg=c(0.5,0,0.5), diff=c(1,0,-1)))
b
```

and doing it step by step

```{r}
coef(a)
crossprod(c(0.5,0,0.5),coef(a))
crossprod(c(1,0,-1),coef(a))
sqrt(crossprod(c(0.5,0,0.5), vcov(a))%*%c(0.5,0,0.5))
sqrt(crossprod(c(1,0,-1), vcov(a))%*%c(1,0,-1))
```

These might be rounded differently when they print, but are the same,eg
```{r}
crossprod(c(0.5,0,0.5),coef(a))-coef(b)[1]
sqrt(crossprod(c(1,0,-1), vcov(a))%*%c(1,0,-1)) - SE(b)[2]
```

And the covariance term:
```{r}
crossprod(c(0.5,0,0.5), vcov(a))%*%c(1,0,-1)
vcov(b)[1,2]
```

You can also use names to indicate coefficients, so that this is the same
```{r}
svycontrast(a, list(avg=c(api00=0.5,api99=0.5), diff=c(api00=1,api99=-1)))
```

### Non-linear

Given a quoted expression in which the free variables are the names of the coefficients, `svycontrast()` treats it as a function $f()$ and computes $f(\hat \theta)$ and $f'(\hat\theta)^T\hat\Xi f'(\hat\theta)$, using `deriv()` to do the symbolic differentiation.

As a trivial case, you can, of course, do linear combinations this way and get the same as above
```{r}
svycontrast(a, list(quote(api00/2+api99/2), quote(api00-api99)))
```
 
Less trivially: geometric means, where $f(\theta)=\exp\theta$ and so $f'(\hat\theta)$ is a diagonal $2\times 2$ matrix with diagonal entries $\exp(\hat\theta)$
```{r}
meanlogs <- svymean(~log(api00)+log(api99), dclus1)
meanlogs
geomeans<-svycontrast(meanlogs,
         list(api00=quote(exp(`log(api00)`)), api99=quote(exp(`log(api99)`))))
exp(coef(meanlogs))-coef(geomeans)

B<-diag(exp(coef(meanlogs)))
crossprod(B, vcov(meanlogs))%*%B - vcov(geomeans)
```    

Even less trivially, `svykappa()` does a bit of quasiquotation as well
```
svycontrast(probs, list(kappa = bquote((.(obs) - .(expect))/(1 - .(expect)))))
```
where `obs` are the names of diagonal entries of a square contingency table and `expect` are the products that define the expected values under independency. For a $2\times 2$ table the expression comes out as
```
(.a_..A_ + .b_..B_ - (.a_ * .A_ + .b_ * .B_))/(1 - (.a_ * .A_ + 
    .b_ * .B_))
```
and `svycontrast()` uses the `deriv()` function to differentiate that expression with respect to all its entries. 