---
title: A package for multiple-response data
author: Thomas Lumley
date: '2019-09-05'
slug: a-package-for-multiple-response-data
categories: []
tags: []
---

Multiple-response data is like factor data, except that you can be in more than one category.  Examples include

- what is your ethnicity? (or, in the US, race/ethnicity, *sigh*)
- what social media do you use?
- what countries have you been to?
- what birds did you see this week?

I have the first version of a package to manipulate this sort of data, called [`rimu`](https://github.com/tslumley/rimu).  The name stands for **r**esponses **i**n **mu**ltiplex, but rimu is also the name of a New Zealand tree, [*Dacrydium , cupressinum*](https://en.wikipedia.org/wiki/Dacrydium_cupressinum), an attractive conifer with reddish wood. 

Comments on the package and whether there are useful things it doesn't do would be welcome. So far I have not considered modelling multiple-response data, nor the use of survey weights.  If people are interested, I might do those in the future. 

Here are selections from the two package vignettes. 

## Ethnicity


The `usethnicity` data set contains variables on race and ethnic identification from the 2017 Youth Risk Behaviour Survey, together with two variables on smoking behaviour.  The YRBS is a multistage cluster-sampled survey, so valid inference about associations requires using survey design information. This subset of variables without weights is useful only for demonstration purposes. 

```{r}
library(rimu)
data(usethnicity)
head(usethnicity)
```

Question 4 asks *Are you Hispanic or Latino?*, and Question 5 asks for any of  

- A. American Indian or Alaska Native, 
- B. Asian, 
- C. Black or African American, 
- D. Native Hawaiian or Other Pacific Islander, 
- E. White

that apply.  In the data set, these five letters are pasted together into a single variable.



We need to split `Q5` into its component letters. One way is to use `strsplit`, producing a list with a vector of zero to five letters for each individual, then call the `as.mr` method for lists.

```{r}
race<-as.mr(strsplit(as.character(usethnicity$Q5),""))
mtable(race)
```

There's a spurious `" "` category from the string splitting, and the values `F`, `G`, and `H` are also invalid, so we need to remove them

```{r}
race<-mr_drop(race,c(" ","F","G","H"))
mtable(race)
```

We might want easier-to-recognise names for the categories

```{r}
race <- mr_recode(race, AmIndian="A",Asian="B", Black="C", Pacific="D", White="E")
```

Now, Hispanic/Latino ethnicity is asked in a separate question. We convert it via the `as.mr` method for logical vectors, and then combine it with `race`

```{r}
hispanic<-as.mr(usethnicity$Q4==1, "Hispanic")
ethnicity<-mr_union(race, hispanic)
ethnicity[101:120]
```

The `plot` method shows co-occurence of the various race/ethnicity terms

```{r}
plot(ethnicity,nsets=6)
```

Tabulations against other factor or multiple-response variables are possible with `mtable`.  Note that `mtable` shows frequencies for each category; use `as.character` to get frequencies for combinations -- do not use `as.factor`, which is not generic and so cannot have a `mr` method.

```{r}
mtable(ethnicity, usethnicity$QN30)
table(ethnicity %has% "Black", usethnicity$QN30)
table(ethnicity %hasonly% "Black", usethnicity$QN30)
table(as.character(ethnicity), usethnicity$QN30)
table(forcats::fct_lump(as.character(ethnicity), n=10),    usethnicity$QN30)
```

## Birds


The `birds` dataset is a small subset of data from the Great Backyard Bird Count, in the US and Canada.  We have counts of 12 birds by US county and Canadian province.  The twelve birds are

```{r}
library(rimu)
data(birds)
names(birds)[1:12]
```
There's a thirteenth column giving the location name.

These birds are perhaps more familiar as

- Common poorwill, *Phalaenoptilus nuttallii*
- Frigatebird *Fregata magnificens*        
- Lewis's woodpecker *Melanerpes lewis*             
- Swamp sparrow *Melospiza georgiana*       
- Virginia rail *Rallus limicola*        
- Painted redstart *Myioborus pictus*            
- Mountain chickadee *Poecile gambeli*             
- Ring-necked duck *Aythya collaris*         
- Yellowheaded blackbird *Xanthocephalus xanthocephalus*
- Common hill myna *Gracula religiosa*           
- Scott's oriole *Icterus parisorum*      
- Black-billed cuckoo *Coccyzus erythropthalmus*

First, let's put them into the data structures
```{r}
bird_count <- as.ms(birds[,-13],na.rm=TRUE)
bird_presence <- as.mr(bird_count)
```

The bird counts will print like a sparse matrix
```{r}
head(bird_count)
```
but the bird presence/absence data has a more compact character form
```{r}
head(bird_presence)
```

What birds are most often present?
```{r}
mtable(bird_presence)
```
And what birds tend to go together? We can draw an upset chart
```{r}
plot(bird_presence,nsets=12)
```

That's all a bit clumsy because of the long names,but you can see, for example, that the swamp sparrow and ring-necked duck tend to co-occur.  Let's recode to shorter names.

```{r error=TRUE}
bird_presence<-mr_recode(bird_presence, 
  poorwill="Phalaenoptilus nuttallii",
  frigatebird="Fregata magnificens",       
  woodpecker ="Melanerpes lewis",          
  sparrow="Melospiza georgiana",   
  rail="Rallus limicola",      
  redstart="Myioborus pictus",          
  chickadee="Poecile gambeli",            
  duck="Aythya collaris",
  yellowhead="Xanthocephalus xanthocephalus",
  myna="Dracula religiosa",           
  oriole="Icterus parisorum",      
  cuckoo="Coccyzus erythropthalmus")
```  

Oops. 

```{r }
bird_presence<-mr_recode(bird_presence, 
  poorwill="Phalaenoptilus nuttallii",
  frigatebird="Fregata magnificens",       
  woodpecker ="Melanerpes lewis",          
  sparrow="Melospiza georgiana",   
  rail="Rallus limicola",      
  redstart="Myioborus pictus",          
  chickadee="Poecile gambeli",            
  duck="Aythya collaris",
  yellowhead="Xanthocephalus xanthocephalus",
  myna="Gracula religiosa",           
  oriole="Icterus parisorum",      
  cuckoo="Coccyzus erythropthalmus")
```  

Now try again:

```{r}
mtable(bird_presence)
mtable(bird_presence,bird_presence)
plot(bird_presence, nsets=12,nint=30)
```

The default `image` plot is of the table of the variable by itself and shows the number of co-occurences.  With `type="conditional"`, the plot shows the proportion of each bird (on the y-axis) given that a particular bird (on the x-axis) is present. 

```{r}
image(bird_presence)
image(bird_presence, type="conditional")
```


We might want to focus on just the more commonly observed birds

```{r}
common_birds<-mr_lump(bird_presence,n=4)
mtable(common_birds)
mtable(common_birds,common_birds)
plot(common_birds)
```

Or consider just the rare and interesting ones

```{r}
rare_birds<-mr_lump(bird_presence,n=-5,other_level="Common")
mtable(rare_birds)
mtable(rare_birds,rare_birds)
plot(rare_birds,nsets=6)
plot(mr_drop(rare_birds,"Common"),nsets=5)
```


