---
title: (What’s up with the brackets?)
author: Thomas Lumley
date: '2019-09-10'
slug: what-s-up-with-the-brackets
categories: []
tags: []
---

In various places, from the R help pages to books to course materials, you see R code like

```{r}
(x = runif(10, 0, 10))
```

which displays the value of `x`. Without the brackets, it doesn't.   Harkanwal Singh, on Twitter, said "I would like to know more". So, in case he's not the only one, this is what's going on.

First, when you evaluate an expression in R it will usually return a value, which will be printed.

```{r}
exp(pi)
rnorm(2)
cos
function(x,y) {
  lm(y~x)
} 
```

The printing doesn't always happen, for example:

```{r}
boxplot(x)
```

Is this because `boxplot` doesn't return a value? No!

```{r}
a<-boxplot(x)
a
```

The value returned by `boxplot` is somehow invisible, but only in a temporary way. In fact, if you look near the end of the `boxplot.default` method, you see

```
if (is.null(pars$boxfill) && is.null(args$boxfill)) 
            pars$boxfill <- col
do.call("bxp", c(list(z, notch = notch, width = width, 
            varwidth = varwidth, log = log, border = border, 
            pars = pars, outline = outline, horizontal = horizontal, 
            add = add, at = at), args[namedargs]))
invisible(z)
```

The value is returned, and tagged as *invisible*.  It won't print, at least not until you do something to it.^[Note: lattice and ggplot objects are different -- for them, the `print` method is what does the plotting]

There are two uses of the invisibility flag that are critical to usability of R but not obvious.  The first is in assignments. If you type something like

```{r}
bigx <- matrix(rnorm(1e6),ncol=100)
```

you **really** don't want `bigx` to automatically print.  But it would, without more effort: these are perfectly good expressions that return a value.

Even worse, you don't want

```{r}
print(x)
```

to print its returned value, because its returned value is typically the value of `x` and that way lies infinite loops. Methods for `print` typically return `invisible(x)`^[even the `print` method for `ggplot` objects, which draws a plot rather than printing]. You **could** also get around this by having `print` not return a value, but that breaks the magic variable `.Last.value` and, anyway, it's not what we do.

And, finally, what's with the parentheses?  The parentheses are a function

```{r}
get("(")
```

It returns its argument unchanged, and, like most functions, it doesn't set the invisibility flag.  So when you type

```{r}
(x = runif(10,0,10))
```
the inner assignment is evaluated, the value is returned invisibly, the value is passed to the `(` function, this returns the same value, but it doesn't set the invisibility flag, and the vector prints. 

My recollection is that this wasn't ever a deliberate plan, and there was a move to get rid of it at some point long ago, which was stopped partly because the trick was being used in so many base R help pages. 