---
title: Hexmaps for NZ District Health Boards
author: Thomas Lumley
date: '2019-11-07'
slug: hexmaps-for-nz-district-health-boards
categories: []
tags: []
---

I'm involved a research project that, among other things, will be comparing various health variables across NZ District Health Boards (DHBs).  In order to make the outputs less boring and (hopefully!) more interpretable, I want some maps.  

This post is about 'DHBins', a set of hexmaps vaguely analogous to the square 'statebins' for US states. The code is in the `DHBins` [package](https://github.com/tslumley/DHBins)  I'll illustrate with some data on immunisation coverage in NZ kids.

First, the data.  It's from the [NZ Social Indicators](http://archive.stats.govt.nz/browse_for_stats/snapshots-of-nz/nz-social-indicators/Home/Health/childhood-immunisation.aspx), giving the percentage of kids turning 5 years old who had completed their age-appropriate vaccinations.  A related set of data is part of the quality targets for DHBs.  We've got data for the whole population and broken down by the broadest ethnicity groups. 


```{r}
library(tidyverse)
library(DHBins)
data(immune)
summary(immune)
```

The `NA` values are when the true values are too small for confidentiality and are suppressed.  First, let's get a picture of the sizes of the DHBs

```{r fig.height=6}
par(mar=c(1,1,1,1))
with(immune,dhbin(radii=sqrt(Ntotal),DHB_names=DHB))
```

Now, the ethnic composition of each one. This is not a pie chart^[The reason this isn't a pie chart is that it's discrete: the difficulty in judging angles in a pie chart doesn't apply. It's a type of glyph map].

```{r fig.height=8}
eth_matrix<-immune %>% 
	select(NNZE,NMaori,NPacific,NAsian,NOther) %>% as.matrix
rownames(eth_matrix)<-immune$DHB
eth_matrix[is.na(eth_matrix)]<-0	
eth_hex<-tri_alloc(eth_matrix, colours=RColorBrewer::brewer.pal(5,"Pastel1"))

dhtri(tri_colours=eth_hex, 
		legend_opts=list(fill=RColorBrewer::brewer.pal(5,"Pastel1"),
		border=NA, legend=c("NZ European","Māori","Pacific","Asian","Other")))
```

The "NZ European" group is much less dominant in 5-year-olds that in middle-aged people; emphasised by this being *prioritised ethnicity*, where multiple ethnicities are reduced to the first in a priority list that goes Māori, Pacific, Asian, Other, NZ European.  The number of triangles a group gets is its proportion of the total rounded to the nearest sixth, with the Sainte-Laguë algorithm used to make this add up to exactly six sixths. 

 

It might be helpful to scale these by population size

```{r fig.height=8}
dhtri(radii=sqrt(immune$Ntotal),tri_colours=eth_hex, 
		legend_opts=list(fill=RColorBrewer::brewer.pal(5,"Pastel1"),
		border=NA, legend=c("NZ European","Māori","Pacific","Asian","Other")))
```

To show the vaccination coverage, we need to code the numeric proportion, taking account of the target being 95% (which is what you need for herd immunity to measles)

```{r fig.height=8}

code_vax<-function(pct,N,dhbnames){
	basecols<-case_when(
		is.na(N)~"lightgray",
		N<75 ~ "lightgray",
		pct< 0.80 ~ "goldenrod",
		pct< 0.85 ~ "red",
		pct <0.90 ~ "orange",
		pct <0.95 ~ "gold",
    	pct>=0.95 ~ "springgreen"
		)
	cols<-adjustcolor(basecols,alpha.f=0.5)	
	names(cols)<-dhbnames
	cols
}

with(immune,dhbin(hex_colours=code_vax(Pcttotal,Ntotal, DHB),
legend_opts=list(fill=c("goldenrod","red","orange","gold","springgreen"),
legend=c("<80%","80-85%","85-90%","90-95%","95%+"), title="Immunisation coverage")
))
```


And now, vaccine coverage by ethnicity; the gray hexes are where the total count is less than 75. 

```{r fig.height=8}

par(mfrow=c(3,2),mar=c(1,0,1,1))
with(immune,dhbin(hex_colours=code_vax(Pcttotal,Ntotal,DHB),short=TRUE))
title(main="All")
with(immune,dhbin(hex_colours=code_vax(PctNZE,NNZE,DHB),short=TRUE))
title(main="NZ European")
with(immune,dhbin(hex_colours=code_vax(PctMaori,NMaori,DHB),short=TRUE))
title(main="Māori")
with(immune,dhbin(hex_colours=code_vax(PctPacific,NPacific,DHB),short=TRUE))
title(main="Pacific")
with(immune,dhbin(hex_colours=code_vax(PctAsian,NAsian,DHB),short=TRUE))
title(main="Asian")
with(immune,dhbin(hex_colours=code_vax(PctOther,NOther, DHB),short=TRUE))
title(main="Other Ethnicities")
```

The short names aren't unique (two Ts, three Ws) but they don't need to be, since no-one who is likely to be looking at the maps will have any chance of confusing Tairawhiti and Taranaki or Waitemata and Waikato.

