---
title: New in the survey package
author: Thomas Lumley
date: '2020-04-03'
slug: new-in-the-survey-package
categories: []
tags: []
---

```{r echo=FALSE, message=FALSE}
library(survey)
data(api)
```

Version 4.0 of the survey package is on its way to CRAN. 

There are two main updates, which improve the estimation of contrasts

First, a couple of improvements to the handling of replicates.  When `svycontrast` is used on an object that includes replicate estimates, the estimates will now be transformed and then used to estimate a variance, rather than using the delta method. I think that's the right  thing to do, though you might also want to compute a confidence interval on the original scale and transform the interval. Also, there's a function to estimate variances from a set of arbitrarily transformed replicates, eg correlation matrices
```{r}
dstrat <- svydesign(id=~1,strata=~stype, weights=~pw, data=apistrat, fpc=~fpc)
bstrat<- as.svrepdesign(dstrat,type="subbootstrap")

v <- svyvar(~api00+api99, bstrat, return.replicates=TRUE)
vcor<-cov2cor(as.matrix(v))[2,1]
vreps<-v$replicates
correps<-apply(vreps,1, function(v) v[2]/sqrt(v[1]*v[4]))

vcov(bstrat,correps, centre=vcor)
```

Second, some functions (so far, `svymean`, `svytotal`, `svyratio`, `svykappa`,`svyglm`, `svymle`) have the option to return influence functions as an attribute of the estimate.  This allows `svyby` to estimate a covariance matrix including between-domain covariances, so that confidence intervals and standard errors are available for between-domain contrasts.  Previously, they have only been available for replicate-weight designs (though for totals and means you could get them with an appropriate regression model). Returning influence functions is not the default, because they can take up a lot of space.

For example

```{r}
dclus1<-svydesign(id=~dnum, weights=~pw, data=apiclus1, fpc=~fpc)

totals<-svyby(~enroll, ~stype, svytotal,
                 design=dclus1, covmat=TRUE)

svycontrast(totals, quote(E/H))
svycontrast(totals, quote(E-H))
```

The only important function where this does not look feasible is `svyquantile`, because of the lack of smoothness (though there are tests for equality of quantiles between groups available with `svyranktest`.)

### Other changes

There is a function `svynls()` to do non-linear least-squares with probability weights.  It does most of its work by calling `nls`, so if you know how to fit your model with `nls` you'll be ok.

There are also a couple of significant documentation notes

- Lots of people have asked how to get 'robust' standard errors for quasipoisson models fitted with `svyglm`, so there is a note saying that there isn't any choice: the standard error estimates are generalisations of the 'sandwich' standard errors
- I had an interesting exchange about standard errors in `predict.svyglm`. I implemented these the obvious way:
$$\mathrm{var}[x\hat\beta]=x^T\mathrm{var}[\hat\beta]x.$$
For the special case where the new value of $x$ is the true population mean, most people use a different standard error estimator that's a rescaling of the weighted residual mean square (especially in simple random samples in textbooks, but also in real life).  If I'm reading Chapter 2 of Fuller's *Sampling Statistics* correctly, my estimator is noisier, but is valid under weaker assumptions. 